---
title: The Urban Fabric as Palimpsest
subtitle: Tracing Proposed Correction Plans in Basel
authors:
  - name: Moritz Twente
    orcid: 0009-0005-7187-9774
    email: moritz.twente@unibas.ch
    affiliation: University of Basel
    corresponding: true
lang: en
keywords:
  - Basel
  - History of Urban Design
  - Automotive City
  - Urban Planning Policies
  - GIS
  - Korrektionsplan
abstract: |
  Like most cities in Switzerland, present-day Basel was shaped particularly by urban planning decisions made between 1940 and 1970: With the underlying goal of modernization, new transport infrastructures were built to make cities more accessible to cars. To this end, several so-called 'Korrektionspläne' were drawn up for Basel that proposed the construction of automobile traffic axes through the Old Town. Overall, these plans were not fully executed, since urban planning that prioritized private cars was met with increasing criticism. Nevertheless, the planned road system left a mark and partial implementations can still be seen today. We use GIS to trace these fragments in the urban fabric. By reading the city structure as a palimpsest to expose the superimposition of temporal layers that each convey different sociopolitical ideas, we critically engage with narratives about the history of Basel.
pagetitle: The Urban Fabric as Palimpsest. Tracing Proposed Correction Plans in Basel.
description-meta: |
  "The Urban Fabric as Palimpsest: Tracing Proposed Correction Plans in Basel" is a paper presentation by Moritz Twente for the conference "Historical Arguments and the Digital", held at the University of Luxembourg in June 2024.
date: 2024-06-25
bibliography: references.bib
#nocite: |
#  @*
citation:
  type: report
  #doi: FIXME
  url: https://mtwente.github.io/palimpsest-bs/
number-sections: true
---

::: {.callout-note}
This document is a work in progress. Check back soon for more content.
:::

Basel prides itself on being a city of rich architectural culture, with its "world-renowned historical city center"^[In the case of Basel, "Historical City Center" refers to both Altstadt Grossbasel on the left bank of the Rhine and Altstadt Kleinbasel on the right.] [@huber_architekturfuhrer_2014] having a front-row seat in marketing brochures, tourist publications, and architecture guides as well. Painted in bright red colour, Basel's City Hall is one example for this type of built heritage. But standing on the market square, visitors might notice the building next to City Hall: The office building at Markplatz 5 (see @fig-marktplatz5mt), built in 1952, stands out against most of the buildings around the square, and makes for an especially stark contrast to the intricatly decorated walls next door.

Like most cities in Switzerland, present-day Basel was shaped particularly by urban planning decisions made between 1940 and 1970 [@eisinger_stadte_2004], which the office building at Marktplatz 5 is a pointed example for – as it opens up a window to an era of Basel urban planning thought to be long gone: Its location at the pedestrianised market square masks the fact that this building owes its very existence to a project with the aim of construction high-capacity roads through the innermost parts of Basel.

In this paper, we will revisit this and similar projects. We will see how different eras of urban planning echo in the urban fabric of contemporary Basel and, focusing on specific planning proposals related to car-centered planning decisions of the mid-20<sup>th</sup> century, how even abandoned plans have left their mark in Basel.

Before delving into any GIS analysis, we will first look at the socio-political circumstances that led to traffic planning becoming a key policy issue in Basel's urban planning at the beginning of the 20<sup>th</sup> century (@sec-context). Then we will quickly dissect the most important aspects of building law and planning regulations (@sec-instruments), before discussing two pre-WW2 plans and their impact (@sec-spiegelhof and @sec-stadthaus). After a change of perspective to built heritage conservation aspects (@sec-heritage) we will then take a look at the available geodata and analyse the overall impact of 20<sup>th</sup> century traffic planning on the contemporary city (@sec-analysis).

::: {.callout-tip}

The sections are associated with individual buildings and streets that, if read from top to bottom, will present you with an alternative city tour of Basel, starting on market square, leading through Altstadt Grossbasel towards Johanniterbrücke, crossing the Rhine into Kleinbasel and back, finishing the tour at Aeschenvorstadt.

:::

# Marktplatz 5: Introduction {#sec-marktplatz5}

Still standing in front of the city hall, we take a closer look at the building at Marktplatz 5. It has a large front window for displaying clothing items to be sold, and a small takeaway stand at its northwestern corner. Comparing today's view with a historical photo taken in 1966 – 14 years after the building was built – reveals that the ground floor, including the front window, was once an open arcade (see @fig-marktplatz5friedli). On its own, this arcade does not have much effect, but the purpose was – in combination with more arcades in the adjacent buildings, to provide a sheltered sidewalk for pedestrians. The space in front of the building that was now freed up was to be converted into an additional lane for motorised traffic and/or parking spaces.

Because the plan to expand the road in front of the building was dropped and the area is now pedestrianised, the building owners decided to use the open arcade space otherwise – one part was incorporated into the store, and the takeaway stand is housed in a container that was placed inside the former arcade.

Just by looking at these two buildings, we can recognize three different eras of urban planning:

- The city hall as remnant of Basel before the turn of the 19<sup>th</sup> century
- The building at Marktplatz 5 with open arcades as result of the focus on car-centered infrastructure since the beginning of the 20<sup>th</sup> century
- The addition of the ground floor extension at a later date as a sign of planning paradigms shifting away from solely car-focused planning.

::: {layout-ncol=2}

![[Marktplatz 5](https://osm.org/go/0CdsJzXoq?m=) to the left of City Hall, facing east. June 2024.](./media/Marktplatz5_20240623_195409.jpg){#fig-marktplatz5mt}

![Marktplatz 5 with open arcades. Detail of a photo by Werner Friedli [-@friedli_basel_1966].](./media/Marktplatz5_ETH-Bibliothek_LBS_H1-026898_Ausschnitt.png){#fig-marktplatz5friedli}

:::

Using the term coined by Corboz [-@corboz_territorium_2001], we can see the city – resp. the built environment – as a palimpsest: Both the city's larger structures and its individual buildings as ever-changing carriers of meaning [-@corboz_geschichte_2001, p. 58], with emphases on different aspects of urban design in each century, each decade.

## The Urban Fabric of the City Center

Marktplatz 5 is only one example of buildings constructed in the last century that illustrates the relatively young age of large parts of the so-called historic city center of Basel. In 1993, Stephan Tramèr made a hand-drawn map comparing the inventory of buildings on an 1865 cadastral plan with the buildings standing at the time – his conclusion: Between 1865 and 1993, around 70% of the urban fabric had been replaced by new buildings [-@meier_basel_1993, p. 10].

A similar comparison is depicted in @lst-map, showing how many of the buildings which today stand in the historic city center(s) were built before 1875.^[I chose to set the comparison between today and 1875 because Tramèr [-@meier_basel_1993, p. 11] based his plan on the basis of the 1875 cadastral plan.]

::: {#lst-map lst-cap="Abrissplan"}

```{r}

library(leaflet)
library(geojsonio)

altstadt1875 <- geojson_sf("data/geodata/buildings1875/altstadt1875.shp")
bezirke_shape <- geojson_sf("data/geodata/districts/100042.shp")

# keep only Altstädte
bezirke_shape <- bezirke_shape[bezirke_shape$wov_label == 12 | bezirke_shape$wov_label == 1, ]

pop_up_buildings <- paste("Constructed in", altstadt1875$GBAUJ)

bs_buildings<-leaflet() %>%
  addTiles()  %>%
  addProviderTiles("CartoDB.Voyager",options = providerTileOptions(
    opacity = 0.7
  )) %>%
  addPolygons(data=bezirke_shape, fillOpacity = 0, weight = 1.5,
              stroke=TRUE, color="black") %>%
    addPolygons(data=altstadt1875, fillOpacity = 0.03,weight = 1.2,
              popup = pop_up_buildings)

bs_buildings

```

:::

A quick glance at this map reveals a spatial clustering: Of the buildings that were part of Basel in 1875, most of the ones still remaining today are clustered in the upper parts of Altstadt Grossbasel and between Greifengasse and Lindenberg in Altstadt Kleinbasel. The clustering becomes even more apparent when compared to a topographic map (@fig-relief) of central Basel:

![Topographic Model of Basel, STA DS BS 9 3370 [@regierungsrat_des_kantons_basel-stadt_ratschlag_1933]](./media/3370_Beilage2_Scan_Topografie.jpg){#fig-relief fig-align="center" width=50% fig-alt="A scanned map showing the topography of Basel."}

Why has so much of the lower part of the old town (also known as _Talstadt_) been torn down? The lower parts of the city center, where the river Birsig used to run before being diverted in a tunnel in 1897 [@vinken_zones_2021, p. 91], house the most important civic and commercial institutions of Basel [@hoflinger_altes_1954, p. 10]. Already at the turn of the century did the narrow valley pose a problem with regard to accessibility. Therefore, the cantonal authorities decided on introducting a large-scale modernisation program for the city center – a Korrektionsplan (e. _correction plan_).

Over the years, more plans of this kind saw the light of day:

```{mermaid}
timeline
  1933 : Schuhmacherplan
  1946 : Correction Plan Kleinbasel
  1949 : Correction Plan Grossbasel
  1958 : Comprehensive Traffic Plan (Leibbrandplan)
  1958 : Counter-Proposal to Leibbrandplan
```

Before taking a look at the content of the first plan, Schuhmacherplan, and its successors, we will expand on the planning problems Basel faced at the beginning of the 20<sup>th</sup> century.

## 20<sup>th</sup> Century Urban Planning in Basel {#sec-context}

As mentioned before, traffic was identified as the main problem to solve by means of the Korrektionsplan. Data provided by the Cantonal Statistical Office [-@statistisches_amt_des_kantons_basel-stadt_statistisches_1951; -@statistisches_amt_des_kantons_basel-stadt_statistisches_1961] shows the number of registered motor vehicles in Basel compared to the numbers of registered motorcycles and registered bicycles, each per 1000 inhabitants (@fig-cars). When cantonal authorities decided on enacting the first plans in the early 1930s, there were barely any motorised vehicles in Basel at all:

```{r}
#| label: fig-cars
#| fig-cap: !expr "Vehicle Types in Basel per 1000 Inhabitants [@statistisches_amt_des_kantons_basel-stadt_statistisches_1951; @statistisches_amt_des_kantons_basel-stadt_statistisches_1961]"
#| column: page-inset-right
#| echo: false
#| message: false


library(ggplot2)
library(tidyverse)

if(!require(plotly)){
    install.packages("plotly",
                     dependencies =TRUE)
    library(plotly)}

vehicles <- read.csv("./data/statistics_cars.csv", sep = ";")

vehicles_long <- vehicles %>%
  pivot_longer(cols = -Year, names_to = "VehicleType", values_to = "Per1000")

carplot <- ggplot(vehicles_long,
               aes(x = Year, y = Per1000, color = VehicleType)) +
  geom_smooth(se = FALSE) +

  scale_x_continuous(
    breaks = seq(1910, 1960, by = 10),
    expand = expansion(mult = c(0, 0))
  ) +

  #geom_point() +
  labs(color = "Vehicle Type",
      xlab = "", ylab = "") +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank())


ggplotly(carplot)
```

Other factors also played roles in the efforts to 'modernise' Basel in the early 20<sup>th</sup> century. As Basel's population grew, a systematic review in [@huber_architekturfuhrer_2014, p. 228] discovered that

- 79% of all flats did not have a bathroom,
- 84% of all flats did not have access to warm water,
- 64% of all flats did not have access to a laundry room.

Additionally, the construction works following the decision to 'modernise' the city center were a welcome job-creation scheme for the local labour market: The correction plan of 1933 played a pivotal role in the _Arbeitsrappen_ job-creation programme that was created in 1936 [@huber_architekturfuhrer_2014, p. 228].

While these additional arguments has to be taken into account when investigating the (first) correction plan(s) of the early 1930s, it is still the topic of traffic planning and allocation of space to the private motor vehicle that fuelled the creation of the plan and its subsequent development in the 1940s.

### Planning Ahead

According to Hans Bernoulli [-@bernoulli_neuer_1933, p. 350], the proposed correction plan

- "In Basel behilft man sich bis auf den heutigen Tag mit den Strassen und Gassen, die für die Verkehrsansprüche des XI. und XII. Jahrhunderts geschaffen worden sind. Kein Wunder, dass eine solche Altstadt pittoresk, aber so gut wie unbrauchbar ist." [@bernoulli_neuer_1933, p. 350]

- "... denn bei Licht besehen, bedeutet solch ein Plan nicht eine Korrektur, sondern einen Neubau der Altstadt ..."[@bernoulli_neuer_1933, p. 350]

Even though private motor vehicles were not a common phenomenon in Basel before World War II (see @fig-cars), they played a pivotal role in the infrastructure planning efforts by the cantonal authorities.

- topography (map with height lines, scan from old document?), Ratschlag 3370
- automobiles (stats: number of cars in Basel?), @fig-cars

* different plans/approaches during 20<sup>th</sup> century:

  - erster Korrektionsplan 1900 [@huber_architekturfuhrer_2014, p. 226]
  - Schuhmacherplan 1930s
  - Korrektionspläne 40/50s => Focus!
  - Leibbrandplan/Gegenvorschlag Fachverbände 60s

* counter narrative of "city center is old" (mostly marketing)
* counter narrative of "Korrektionspläne nur ansatzweise umgesetzt"

  - [@fingerhuth_bauten_1988, p. 50], [@fingerhuth_bauten_1988, p. 52]
  - "Die Ansätze [!] zur Verwirklichung des Plans zeigen sich besonders einschneidend im Bereich Blumenrain-Spiegelgasse, am Rümelinsplatz und am Barfüsserplatz" [@huber_architekturfuhrer_2014, p. 227]
  - “Besonders in der Talstadt mussten im 20. Jahrhundert viele Häuser dem Konzept der sogenannten Talentlastungsstrasse weichen, die vom Blumenrain bis zum Barfüsserplatz hätte führen sollen. Die Neuausrichtung und Verbreiterung der Spiegelgasse mit der Kantonalbank und dem Spiegelhof (erbaut 1937–1939) machten hiermit den Anfang, es folgten Neubauten in der Stadthaus-, der Schneider- und der Münzgasse. Erst 1974, ein Jahr vor dem europaweiten Denkmalschutzjahr, wurden die Verbreiterungspläne aufgegeben.” [@mohle_profanes_2016, p. 103]
  - "Dank dem Abbruch des Vorhabens blieb das Gerbergässli verschont und die verbreiterte Strasse fand am südlichen Platzende ein abruptes Ende. Noch heute mahnt das auffällige Eck beim Rümelinsplatz 1 daran, dass Basel nur knapp der Zerstörung seiner Altstadt entgangen ist." [@russi_basler_2022, p. 33]
  - "zum Fischmarkt: “Die neuen Zweckbauten sowie die überaus breite Spiegelgasse – Teil einer geplanten, aber nie finalisierten Talentlastungsstrasse – beraubten den Ort nicht nur seiner Ästhetik, sondern auch seiner Nutzfläche.” [@russi_basler_2022, p. 62]
  - "Ab den 1930er-Jahren wurde das Auto zum Planungsfaktor, sichtbar in den Fragmenten der geplanten Talentlastungsstrasse am Blumenrain um an Fischmarkt, in der Erneuerung der Steinentorstrasse und im Abbruch der alten Aeschenvorstadt" [@zaugg_stadtentwicklung_2010, p. 160]

  dagegen: Schuhmacherplan prägte Stadtplanung bis in die 1960er-Jahre [@huber_architekturfuhrer_2014, p. 226]

# Rümelinsplatz: Talentlastungsstrasse

- idea of Talentlastungsstrasse (see topography)
- how to implement?

::: {layout-ncol=2}
![[Rümelinsplatz 1](https://osm.org/go/0CdsJnGmj?m=), facing north. June 2024.](./media/Ruemelinsplatz1_20240623_142351.jpg)

![[Rümelinsplatz 5](https://osm.org/go/0CdsJnGOI?m=), facing south. June 2024.](./media/Ruemelinsplatz5_20240623_142303.jpg)
:::

## Planning Instruments {#sec-instruments}

- introduction into planning regulation

  Korrektionsplan, Ratschlag 4531:

### Korrektionsplan

Correction Plans were used as strategic planning documents on a large scale. Adressing traffic engineering solutions was one, if not the main objective, as stated during the plan's development by the cantonal parliamentarians [-@grosser_rat_des_kantons_basel-stadt_bericht_1949, p. 5]:

> The correction plan aims to:<br /> 1. relieve the city centre of all unnecessary traffic by expanding well-dimensioned and well-located bypass roads,<br />2. improve urban traffic itself by expanding and supplementing the existing road network,<br />3. reorganise conditions that have become untenable.

How does it work?

- propose new perimeter block outlines, essentially determining road space
- propose arcades

::: {layout-ncol=2}

![Correction Plan Proposal for Grossbasel (Detail), Ratschlag 4531 [-@grosser_rat_des_kantons_basel-stadt_bericht_1949]](./media/R4531_Korrektionsplan_Ausschnitt.jpg)

![Correction Plan Proposal for Kleinbasel (Detail), Ratschlag 3370 [-@grosser_rat_des_kantons_basel-stadt_ratschlag_1933]](./media/R3370_Korrektionsplan_Ausschnitt.png)

:::

The scope of a correction plan was on a city-wide level: To provide a long-term [@regierungsrat_des_kantons_basel-stadt_ratschlag_1933, p. 26], integrated future framework of reference for city development (solely regarding traffic planning, that is). While not addressing individual plots or buildings per se, it is a document operating on a larger scale to show the overall planning objective that was agreed upon. To actually enact the planning decisions that are depicted in the correction plan, the planning authorities then have to transfer the planning objectives to a smaller scale using _Baulinienpläne_.

### Baulinienplan

![Map regarding Baulinienplan 4977 [Detail, -@regierungsrat_des_kantons_basel-stadt_baulinienplan_1960], Location: [Hammerstr./Feldbergstr.](https://osm.org/go/0CdsPMmYk--?m=)](./media/Baulinienplan4977_Ausschnitt.png)

The map, a legal document, is accompanied by a textual resolution with further details on the plan. It includes, among other things, numerical representations of the changes depicted on the map and a list of the plots that are affected by this plan (including information on each plot's ownership status). This document is then legally binding for authorities and plot owners alike

::: {layout-ncol=2}

![Resolution 1818 regarding Baulinienplan 4977, Excerpt. [-@regierungsrat_des_kantons_basel-stadt_beschluss_1960]](./media/Beschl_000986_Ausschnitt_p1.jpg){width=50%}

![Resolution 1818 regarding Baulinienplan 4977, Excerpt. [-@regierungsrat_des_kantons_basel-stadt_beschluss_1960]](./media/Beschl_000986_Ausschnitt_p2.jpg){width=50%}

:::

...

# Spiegelhof: Schuhmacherplan {#sec-spiegelhof}

- 1930s
- jump on scale
- criticism?

  - Schuhmacher selbst: "Im Ratschlag ist ausdrücklich darauf hingewiesen, dass die Altstadt verhältnismässig viel Verkehrsfläche aufweist." [@bernoulli_neuer_1933, p. 352]

- Plan gemindert vom grossen Rat, trotzdem finanzielle Überforderung [@huber_architekturfuhrer_2014, p. 227]

::: {layout-ncol=2}

![[Spiegelgasse 6](https://osm.org/go/0CdsMg8AS?m=), facing south. June 2024.](./media/Spiegelgasse6_20240623_141146.jpg)

![[Blumenrain 25](https://osm.org/go/0CdsMOvf6?m=), facing southwest. June 2024.](./media/Blumenrain25_20240623_135654.jpg)

:::

# Schneidergasse 18: Stadthaus {#sec-stadthaus}

- Scope of 1940s correction plans
- doubts about feasibility
- Ziele klar: "Ziele waren klar die Vervielfachung der Verkaufs- und Bürofläche und die Zugangserleichterungen für den Individualverkehr (Grossparkings in der Innenstadt: ‘Storchen’ am Fischmarkt, ‘Autosilo’ an der Steinentorstrasse, ‘Drachen’ in der Aeschenvorstadt)." [@huber_architekturfuhrer_2014, p. 355]
  - Feststellung, dass man auch parken muss
  - neues Argument: Geschäftsflächen
- neu Arkaden, teilweise Erhaltung

::: {#lst-correctionlines lst-cap="Correction Lines as proposed by the 1949 plan for Grossbasel"}

```{r map-correction-lines}

library(leaflet)
library(geojsonio)
library(tidyr)

lines3370r_bpbs_844_33 <- geojson_sf("data/geodata/correction_lines/3370r_bpbs_844_33_linien.shp")
lines3433b_kpbs_1395_34 <- geojson_sf("data/geodata/correction_lines/3433b_kpbs_1395_34_linien.shp")
lines4172r_kpkb_5100_45 <- geojson_sf("data/geodata/correction_lines/4172r_kpkb_5100_45_linien.shp")
lines4224r_kpgb_4100_46 <- geojson_sf("data/geodata/correction_lines/4224r_kpgb_5100_46_linien.shp")
lines4237_kpkb_5400_46 <- geojson_sf("data/geodata/correction_lines/4237_kpkb_5400_46_linien.shp")
lines4531b_kpgb_4500_49 <- geojson_sf("data/geodata/correction_lines/4531b_kpgb_4500_49_linien.shp")

map4531b<-leaflet() %>%
  addTiles()  %>%
  addProviderTiles(providers$SwissFederalGeoportal.SWISSIMAGE, group = "Swisstopo Satellite Imagery") %>%
  addProviderTiles("CartoDB.Voyager", group = "OpenStreetMap (CartoDB)") %>%

    addPolylines(data = lines3370r_bpbs_844_33, fillOpacity = 1, weight = 3, color = "#1965B0",
                group = "Proposal 3370 for Basel (1933)") %>%
    addPolylines(data = lines3433b_kpbs_1395_34, fillOpacity = 1, weight = 3, color = "#7BAFDE",
                group = "Proposal 3433 for Basel (1934)") %>%
    addPolylines(data = lines4172r_kpkb_5100_45, fillOpacity = 1, weight = 3, color = "#F6C141",
                group = "Proposal 4172 for Kleinbasel (1945)") %>%
    addPolylines(data = lines4224r_kpgb_4100_46, fillOpacity = 1, weight = 3, color = "#F1932D",
                group = "Proposal 4224 for Grossbasel (1946)") %>%
    addPolylines(data = lines4237_kpkb_5400_46, fillOpacity = 1, weight = 3, color = "#E8601C",
                group = "Proposal 4237 for Kleinbasel (1946)") %>%
    addPolylines(data = lines4531b_kpgb_4500_49, fillOpacity = 1, weight = 3, color = "#DC050C",
                group = "Proposal 4531 for Grossbasel (1949)") %>%

    addLayersControl(
      baseGroups = c("OpenStreetMap (CartoDB)", "Swisstopo Satellite Imagery"),
      overlayGroups = c("Proposal 3370 for Basel (1933)", "Proposal 3433 for Basel (1934)",
                      "Proposal 4172 for Kleinbasel (1945)", "Proposal 4224 for Grossbasel (1946)",
                      "Proposal 4237 for Kleinbasel (1946)", "Proposal 4531b for Grossbasel (1949)"),
      options = layersControlOptions(collapsed = FALSE)
    )

map4531b
```

:::

# Petersgraben 20: Heritage Conservation {#sec-heritage}

...
[@huber_architekturfuhrer_2014, p. 227]

- Altstadtschutzzone 1939, unterschiedliche Bedeutungen von Heritage Conservation
  "Strictly speaking, Basel's famed Altstadt came into existence as a planning zone" [@vinken_zones_2021, p. 91]

- "Die Furcht der Bevölkerung vor dem Verlust des vertrauten Stadtbildes zeigt sich auch im zunehmenden Einfluss der Heimatschutzbewegung. Der Abbruch des Zeughauses (heute Kollegienhaus), des Segerhofes am Blumenrain und des Württembergerhofes (heute Kunstmuseum) setzte in den mittleren 1930er-Jahren ein Alarmzeichen. Die denkmalpflegerischen Massnahmen dieser Zeit orientierten sich eher an einer allgemeinen, bildhaften und auch motivischen Nähe zum Altbau denn an den heute vertretenen analytischen Grundsätzen" [@huber_architekturfuhrer_2014, p. 229]

![1939 Zoning Map of Basel. The the area shaded in purple represents the Schutzzone Altstadt [Detail, -@regierungsrat_des_kantons_basel-stadt_ratschlag_1939]](./media/R3769_Zonenplan_39_Schutzzone.jpg)

- "Diesen trug das neue Hochbautengesetz von 1939 in einem gewissen Sinne Rechnung mit der sog. Altstadtzone; in dieser sollte die ‘architektonische Struktur’ (nicht aber die materielle und baukünstlerische Substanz der Häuser) erhalten bleiben. Die in einem Verzeichnis aufgeführten ausgewählten Baudenkmäler durften demnach erst abgebrochen werden, wenn die Stadt deren Kauf ablehnte" [@huber_architekturfuhrer_2014, p. 227]

- "Bereits 1939 waren die Hügelrücken nahe der Peterskirche und der Leonhardskirche als Altstadtzonen ausgewiesen worden und unterlagen seither einem besonderen Baurecht. Seitdem wurden bei denkmalpflegerisch begleiteten Umbauten fast in jedem Haus historische Konstruktionen und Dekorationsschichten freigelegt und restauriert oder zumindest gesichert.” [@mohle_profanes_2016, p. 103]

- Ratschlag 5468: Aufhebung Baulinien Spalengraben, "Im Bestreben, schöne Partien unserer Altstadt zu erhalten, …": Grabeneck zurückverlegt 19.10.1950, jetzt auch Hs. 18–24

  - "Im Korrektionsplan aus dem Jahre 1934 wurde die Südseite des Spalengrabens mit einer Korrektionslinie belastet. Es war vorgesehen, die alte Gebäudefront abzubrechen und den Südflügel des Kollegiengebäudes der Universität freizugeben. In der Zwischenzeit ist der Sinn für die Schönheit unserer Altstadt gewachsen; es wird Wert auf deren Erhaltung gelegt."

  - 6490: "Der Vollständigkeit halber sei hier erwähnt, dass der Altstadtschutz nicht mit dem Denkmalschutz verwechselt werden darf. Hier handelt es sich um eine Bewahrung des überlieferten Stadtbildes durch besonders strenge bauliche Vorschriften, während dort einzelne Gebäude unter Schutz gestellt, d.h. praktisch mit einem Abbruchverbot belegt werden."

  - "The Schutzzonen Altstadt, however, were, although they were designated as protection zones, were [sic] likewise planning zones, and conservation legislation in the proper sense was not established until 1977" [@vinken_zones_2021, p. 92]

  Kostenaspekt der Unterschutzstellung: [@kreis_schicksal_2005, p. 54]

  spezifisch Aeschenvorstadt: "Die Vorbehalte gegen die von den Behörden vorgespurte Erweiterung der Aeschen hatten in den vier Jahren zwischen 1949 und 1953 stark zugenommen. Dem stand allerdings eine gewisse Schlussstrichstimmung entgegen. 1949 hatten noch alle grossen Parteien den Korrektionsplan gutgeheissen. 1953 waren nur noch die Radikalen und Sozialdemokraten für die Durchsetzung der ursprünglichen Ordnung, die Liberalen gaben die Nein-Parole heraus, und die übrigen (die Katholische Volkspartei, die Evangelische Volkspartei, die Bürger- und Gewerbepartei, der Landesring und die PdA) beschlossen Stimmfreigabe. Der Vorstand des Heimatschutzes unterstütze 1953 das Komitee, 1949 hatte er aus Rücksicht auf die internen Befürworter des Korrektionsplanes auf eine Ja-Parole verzichtet und das Komitee nur nachträglich finanziell unterstützt." [@kreis_schicksal_2005, p. 57]

# Spitalstrasse: Unhappy Doctors

- Ratschlag 3370: Freiwilligkeit?

- Kreis 2005, p. 54: "für die öffentliche Hand kostengünstige Zurückversetzung der Baulinie auf der rechten Seite vor, da die meisten Eigentümer im Hinblick auf ihre grösseren Renditen bereit waren, den Boden für die zustäzliche Allmend dem Staat kostenlos abzugeben"

- who is against having cars everywhere? Hospitals

- Referendumsdemokratie, Abstimmungen 1949

  - emotional debate, show two ads each (Goebbels/communism and 'ebi patria' + x)
  - achtung: die Schweiz etc.
  - comparison of city to growth of children, Plakat
  - Aktion Aesche 1950s: "Zugleich wurde ihnen [den Gegnern] aber die Sachkompetenz abgesprochen, die Opposition lasse sich, hiess es, vor allem von Gefühlen leiten." (Kreis 2005, 59)

- Abstimmungsergebnisse als Diagramm

### Results per Polling Place^[without Military and 'Sick'/'Frail' votes]

```{r fig-referendum}
library(tidyr)

data <- read.csv("./data/1949_referendum_results.csv")

data$yes_pct <- data$yes_pct/100
data$no_pct <- data$no_pct/100

# remove Militär and Kranke und Gebrechliche
data <- data[-24:-25,]

# Convert data to long format for ggplot
data_long <- data %>%
  pivot_longer(cols = c(yes_pct, no_pct),
               names_to = "Vote",
               values_to = "Percentage")

# Create the stacked bar chart
plot1949 <- ggplot(data_long, aes(x = polling_place, y = Percentage, fill = Vote)) +
  geom_bar(stat = "identity", orientation = "Percentage") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black") +
  theme_minimal() +
  #coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("yes_pct" = "#6195CF", "no_pct" = "#A5170E")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "", y = "")

ggplotly(plot1949)
```

# Feldbergstrasse: City-Wide Implementation?

```{r map-Feldbergstr31}

if(!require(leaflet)){
    install.packages("leaflet",
                     dependencies =TRUE)
    library(leaflet)}

leaflet() %>%
  addProviderTiles(providers$SwissFederalGeoportal.SWISSIMAGE) %>%
  setView(7.58983, 47.56624, zoom = 20) %>%
  addMarkers(lng=7.58983, lat=47.56624, popup="Feldbergstrasse 31, 4057 Basel")
```

- playing the long game: nobody expects quick results

  - Bürgergemeinde inquiry on renovation timeframes
  - Schneidergasse 11: https://oereblex.bs.ch/api/geolinks/3574.html (Betting on Red Wine), Rücknahme 1978 https://oereblex.bs.ch/api/geolinks/2554.html

- Verlangsamungsgründe:

  - 2. Weltkrieg
  - "Der Ausbruch des Zweiten Weltkriegs verhinderte zunächst eine Weiterprojektierung. Danach waren es die ernüchternden Erkenntnisse aus der Umsetzung des Korrektionsplans, die lediglich partiell erfolgt war, was zu einem vorläufigen Verzicht führte. Insbesondere der hohe finanzielle Aufwand und die Belastung der ansässigen Bevölkerung durch die umfangreichen und langanhaltenden Bauarbeiten wurden als kritisch beurteilt; aber auch ein zunehmendes denkmalschützerisches Bewusstsein stellte solch einschneidende Vorhaben in Frage" [@lenherr_fortwahrendes_2022, p. 92]

  - Kleinbasel plan was already put in place earlier, resp. it took longer to finish the Grossbasel plan, reasons given in Ratschlag 4172: "Kleinbasel bietet einer Korrektion wegen seines klaren Aufbaues weniger Schwierigkeiten als Grossbasel. Die vorhandenen Hauptstrassen liegen alle zweckmässig. Auf Durchbrüche oder neue Strassen kann infolgedessen vollständig verzichtet werden. Es genügt eine allmähliche, durch Baulinien zu sichernde Anpassung der vorhandenen Strassen an die Erfordernisse des zukünftigen Verkehrs. Dabei ist es auch möglich, den mittelalterlichen Kern, soweit er nicht sanierungsbedürftig ist, weitgehend zu erhalten." (p. 1)

- Kostenfrage: "Because of the size of the compensation payments to owners, this process was only economically and politically viable for a few main shopping streets." [@vinken_zones_2021, p. 91]

- Kreis 2005: “Ein weiteres wichtiges Element dieser ‘Grosskorrektion’ sah eine für die öffentliche Hand kostengünstige Zurückversetzung der Baulinie auf der rechten Seite vor, da die meisten Eigentümer im Hinblick auf ihre grösseren Renditen bereit waren, den Boden für die zustäzliche Allmend dem Staat kostenlos abzugeben. Die privaten Investoren (insbesondere der Briefmarkenhändler Ernst Müller-Price mit dem Architekturbüro Bräuning, Leu und Dürig sowie die Schweizerische Kreditanstalt) waren für die untere Aeschenvorstadt die treibenden Kräfte und stiessen bei den Behörden aus dem genannten Grund auf offene Ohren. p. 54

– Ratschlag 4172, Feldbergstrasse: 1934 Verbreiterung zw. Claragraben u. Riehenring vorgesehen auf 18m, 4172 schlägt vor: 20m mit Fahrbahnbreite 12m. (p. 8)

## GIS Analysis {#sec-analysis}

As described above, the correction plans and their impact barely play a role in narratives about Basel's city center. On the contrary: Publications that do mention the plans largely downplay the degree of implementation and focus on the Blumenrain–Rümelinsplatz axis. The aim of this paper is to challenge this notion and to look for traces of these proposals in the historical city center and beyond. To this end, we make use of several data sets that are available with information on Basel's urban fabric today.

### Data Sources and Description

... mermaid graphs

### Calculating Distances Between Buildings and Streets

The planning approach proposed by the correction plans' authors was to make more space available for vehicles, i.e. to widen those roads that were deemed necessary to guarantee a steady flow of (car) traffic. The cheapest way to achieve this – in the already built-up districts of the inner city and closer suburbs – was essentially to move the Baulinie further away from the existing roads [@kreis_schicksal_2005, p. 54] in order to free up space for additional lanes and/or parking. Therefore, investigating the degree of the correction plans' implementation essentially boils down to looking for roads that are wider than others.

But road width alone is no feasible indicator for this research task: As pointed out by the literature, the correction plans never saw _full_ implementation, which means that only in a very few places – e.g. at Petersgraben or Aeschenvorstadt – we actually see roads that were widened. And since the automotive city has been retired as a planning paradigm by now, it is even less likely that any additional space accommodated for through traffic is actually used as such. Instead, the lanes have been kept at their original width and the additional space is used for bicycle stands, trash collection containers (at Petersgraben), outdoor seating for restaurants (at Aeschenvorstadt), etc.

Instead, we have to rely on other variables. The distance between the roads and adjacent buildings appears to be a second indicator: At places that saw partial implementation of the correction plans, we can see that even if a road was not expanded with an additional lane, buildings might have been placed at a distance to the road that accommodated for planned expansion of the road that building was built next to.

The first steps of the analysis are therefore to calculate the width of each road in Basel, and then calculating distances between each building and the road(s) in the vicinity of it.

In the following sections, I am going to describe the approach to calculating these attributes in QGIS and the data that was used for that purpose.

### Calculating Road Widths

*check mermaid shapes: source DB, polygons: stadium, line features: square, numerical features: [[]], output {{}}
*colours: input D9CCE3, intermediate results 7BAFDE, output F7F056

```{mermaid fig-WorkflowWidth}
flowchart TD
    RN4 --> RW1
    subgraph Road Network
        RN1[(City Boundary)] -->|Download from OSM| RN3[Basel Transport Network]
        RN2[(Transport Network)] -->|Download from Swisstopo| RN3
        RN3 --> |Filter Features| RN4[Basel Road Network]
    end

    RN4 -->|Calculate Bearings| RW2

    subgraph Road Area Geometries
        RG1[(Ground Cover Data)] -->|Download from Canton| RG2([Transport Geometries])
        RG2 --> |Filter Features| RG3([Way Area Polygons])
        RG3 --> |Fix and Dissolve| RG3
    end

    RG3 --> |Transform to Lines| TL1
    RG3 --> TL3

    subgraph Transect Lines
        TL1[Way Area Outlines] -->|Transect 25m| TL2[Perpendicular Lines]
        TL2 -->|Clip to Area Polygons| TL3[Lines Representing Road Width]
    end

    TL3 -->|Filter Excess Roads| RW1

    subgraph Road Width
        RW1[Transects within Road Network] --> RW3
        RW1 --> |Calculate Bearings| RW2
        RW2[[Angle Difference]] -->|Filter Right Angles| RW3{{Road Width Line Features}}
        RN4 --> |Filter Geometry Outliers| RW3
    end

  style RN1 fill:#D9CCE3
  style RN2 fill:#D9CCE3
  style RG1 fill:#D9CCE3
  style TL3 fill:#7BAFDE
  style RN4 fill:#7BAFDE
  style RG3 fill:#7BAFDE
  style RW3 fill:#F7F056

```

### Calculating Distance To Road

```{mermaid fig-WorkflowDistance}
flowchart TD

    subgraph Building Footprints
        BF1[(Building Data)] -->|Download from Canton| BF2([Building Footprints])
        BF2 --> |Transform to Outline| BF3["Building Outlines (Façades)"]
        BF3 --> |Calculate Centroids| BF4(Façade Centroids)
    end

    subgraph Road Areas
        RA1[(Ground Cover Data)] --> |Download from Canton| RA2([Transport Geometries])
        RA2 --> |Filter Features| RA3([Way Area Polygons])
        RA3 --> |"Buffer -0.01m"| RA3
        RA3 --> |Transform to Lines| RA4([Way Area Outlines])
    end

    RA4 --> BF5

    subgraph Distance Calculation
        BF4 --> |Draw Shortest Lines| BF5[Shortest Line from Façades to Road Area]
        BF5 --> |Split At Intersections| BF6[Line Segments Outside of Building]
        BF3 --> |Difference| BF6
        BF6 --> |Keep Only Intersects with Road | BF6
        BF6 --> |Rank Line Lengths| BF7[Shortest Distance per Building]
        BF7 --> |Extract Vertices| BF8(Lines Originating From Street-Facing Buildings Only)
    end

    subgraph City Blocks
        CB1[(Statistical City Blocks)] --> |Download from Canton| CB3([Perimeter Blocks])
        CB2[(City Boundary)] --> |Download from OSM| CB3
        CB3 --> |Buffer In -20m| CB3
        CB3 --> |Buffer Out 1.5m| CB3
    end

    CB3 --> |Filter Features| BF8

    subgraph Building Statistics
        BS1[(Building Statistics)] --> |Download from Canton| BS2[[Building Type, Age etc.]]
    end

    BS2 --> |Join Attributes| BF8

    subgraph Clean Outliers
        CO1[[Building Distances]] --> |Calculate Quartile Ranges| CO2[[Upper/Lower Bounds]]
        CO2 --> |Filter Outliers| CO3{{Building/Street Distance Line Features}}
    end

    BF8 --> CO1

  style BF1 fill:#D9CCE3
  style RA1 fill:#D9CCE3
  style CB1 fill:#D9CCE3
  style CB2 fill:#D9CCE3
  style BS1 fill:#D9CCE3
  style BF4 fill:#7BAFDE
  style RA4 fill:#7BAFDE
  style BF8 fill:#7BAFDE
  style CB3 fill:#7BAFDE
  style BS2 fill:#7BAFDE
  style CO3 fill:#F7F056

```

# Aeschenvorstadt 68: Last Man Standing {#sec-aeschenvorstadt}

The Aeschenvorstadt is arguably the last street where measures proposed by the correction plans were implemented. Plot construction plans on the basis of the 1940s correction plans were still being issued, when in 1953 a group of citizens managed to launch a contested referendum on the correction plan that propsed the demolition of the western side of the road [@suter_von_1991, p. 25]. And even though reservations against the road expansion at Aeschenvorstadt had increased considerably in the years before the referendum and most political parties did not support the implementation of the road expansion anymore [@kreis_schicksal_2005, p. 57], the voters^[Only men at the time: In Basel-Stadt, cantonal voting rights for women were introduced in 1966 [@crain_merz_grosse_2021]], narrowly approved the plans, essentially paving the road for the demolition of Aeschenvorstadt.

- Aeschenvorstadt: last implemented policies from Korrektionsplan
  - why so late still? Kreis 2005 zu **Eigentümerstruktur**: "Vielleicht waren für diese Variante [Alternativorschlag anstelle Aeschenverbreiterung] die in solchen Fällen stets sehr wichtigen Besitzverhältnisse weniger günstig, das heisst: Der Besitz der Aeschenparzellen lag in Händen von Leuten, die über grosse Nähe zu den Behörden verfügten." (p. 58), s. auch:
    - Das Gebiet um den Andreasplatz blieb ausgespart und war 1949 Gegenstand eines Wettbewerbs mit dem Ziel, den Marktplatz an die Schneidergasse zu verlegen (1. Preis: Bräunig, Leu, Dürig). [@huber_architekturfuhrer_2014, p. 229]

“Im September 1953 wurde per Volksabstimmung die Opferung historischer Bausubstanz der Aeschenvorstadt (‘Goldener Sternen’) mit einem Ja von 59,7 Prozent zum schliesslich nie zu Ende geführten Korrektionsplan angenommen.” [@kreis_goldene_2000, p. 276]

"Daher wurde ein Korrektionsplan ausgearbeitet, der auf der westlichen Vorstadtseite eine beträchtliche Zurücksetzung der Bau- und Straßenlinie und ihre Begradigung vom Banken- bis zum Aeschenplatz vorsah. Gegen diesen von der Regierung vorgelegten und vom Großen Rat gutgeheißenen Korrektionsplan wurde das Referendum ergriffen. Nach einem heftigen Abstimmungskampf unterlagen indes die dem Heimatschutzgedanken verpflichteten Gegner im September 1953 mit bloß 350 Stimmen. Damit war der Untergang einer noch fast intakten malerischen Altstadtstraße besiegelt […]" [@suter_von_1991, p. 25]

“Bei der Schleifung der beiden letztgenannten meinten die seither ebenfalls untergegangenen ‘Basler Nachrichten’: ‘Kommende Generationen werden dieses Zerstörungswerk nie begreifen können.” [@suter_von_1991, p. 25]

tides are turning:

"Selbst Sanierer - vielleicht gerade sie – führten das Wort von den 'notwendigen Opfern' im Munde. Weil man bei anderen Teilen der Altstadt für deren Erhaltung eintrat, glaubte man in diesem Fall das 'Opfer' den Freunden des alten Basel zumuten zu können." (Kreis 2005, p. 58)

“Anfang der siebziger Jahre wurde das Verkehrskonzept für die Innerstadt neu überdacht. Die Baulinien für die Talentlastungsstrasse im Raum der Schneidergasse wurde 1974 aufgehoben. Auf die totale Auskernung im Bereich Schneidergasse-Spalenberg-Nadelberg wurde verzichtet. Noch immer bestand aber die Vorstellung von der radikalen Sanierung der Altstadt.” [@fingerhuth_bauten_1988, p. 52].

"1964 sprach sich die Regierung für das Konzept des sogenannten Cityrings aus und erteilte der Vorstellung, dass der Durchgangsverkehr durch das Stadtzentrum führen müsse, eine definitive Absage. Damit wurde auch das in der Zwischenkriegszeit entwickelte, im Korrektionsplan von 1947 bestätigte und bis Ende der 1950er-Jahre gültige Konzept begraben, das eine zusätzliche Transitstrasse durch die Schneidergasse und das hintere Gerbergässlein (so genannte Talentlastung) vorgesehen hatte.” [@kreis_goldene_2000, p. 276]

"Nach dem Ölschock der 1970er-Jahre zeichnete sich ein Bewusstseinswandel ab. Die Bewohner wollten keine neue Stadt mehr, sie wollten die alte zurückhaben." [@zaugg_stadtentwicklung_2010, p. 160]

"Bis in die späten 1960er Jahre war der Einsatz für den Schutz, wie ihn die Aeschenvorstadt nötig gehabt hätte, eine Sache der Minderheit, die inzwischen tendenziell zur Mehrheit geworden ist. Das grosse Umdenken fand, wie Uta Feldges richtig feststellt, erst in den 1970er Jahren statt. Es wurde von der politischen Linken mitgetragen, die in den 1930er wie in den 1950er Jahren ein kräftiger Motor der zerstörerischen Modernisierung war." (Kreis 2005, p. 59)

"Erst 1974, ein Jahr vor dem europaweiten Denkmalschutzjahr, wurden die Verbreiterungspläne aufgegeben" [@mohle_profanes_2016, p. 103] – _spezifisch bzgl. Talentlastung?_

## new times

- Rolling back the regulations: starting ca. 1970s?

  - Klybeckstr. 1934 https://oereblex.bs.ch/api/geolinks/2948.html, 1993 https://oereblex.bs.ch/api/geolinks/2947.html
  - Hardstr. 1935 https://oereblex.bs.ch/api/geolinks/2798.html, 2022 https://oereblex.bs.ch/api/geolinks/4941.html (Street Line removed, Building line still existing)

- Pharmacy Aeschenvorstadt

  - large number of individual plot owners: https://oereblex.bs.ch/api/geolinks/3505.html

- other trends in urban planning gained traction: Cityring
  - problem shifts from transit/accessibility to providing parking:
    "Nun entwickelte man auch neue Lösungen für die Parkprobleme. Die 1956/60 erbaute Storchen-Garage (die ursprünglich unterirdisch bis zum Spalentor hätte ausgedehnt werden sollen), ist ein anachronistisches Relikt aus der Vor-Cityring-Ära. Das 1956/58 erbaute Autosilo an der Steinentorstrasse (es war das erste vollautomatisierte Parkhaus der Welt) wurde wieder aufgehoben, als von 1968 an in den Fundamenten des Heuwaage-Viaduktes neue Abstellmöglichkeiten zur Verfügung standen. Das erste Stück einer Fussgänger-City wurde jedoch erst 1972 geschaffen" [@kreis_goldene_2000, p. 276]
  - “Bereits 50 Jahre später stand die nächste Umgestaltung an, denn nun verfolgten die Behörden ihren Plan einer Autobahn mitten durch die Innenstadt. Das vom ETH-Ingenieur – und Nazi-Kriegsverbrecher – Kurt Leibbrand entwickelte Projekt sollte den Stadtkern entlasten, in dem die angrenzenden Parallelstrassen zu mehrspurigen, möglichst gerade verlaufenden Verkehrsadern ausgebaut werden [@russi_basler_2022, p. 33]
  - "1962 stellte ETH-Ingenieur Kurt Leibbrand, auch «Karajan des Verkehrswesens» genannt, den Gesamtverkehrsplan Basel vor. Die Idee: Einen Ring um den Stadtkern herum bauen, der den Verkehr aufsaugt." [@wieland_platz_2022]
  - “Die Diskussionen um den neuen Viadukt wurden bereits vor der Realisierung, vor allem aber danach kontrovers geführt. Aus verkehrsplanerischer Sicht war er zunächst vielleicht ein Gewinn. Städtebaulich stellte er hingegen einen markanten Einschnitt dar und schuf Probleme, die erst im Laufe der Zeit in vollem Umfang sichtbar wurden. So prangerte das ‘Aktionskomitee gegen den Cityring und für ein wohnliches Basel’ das ‘scheussliche Stahlmonument’ an, für das die Elisabethen- und Steinenschanze gerodet werden musste. Insbesondere die Fussverbindung von der Altstadt zum Bahnhof und zum Zoo war wegen der komplexen Verkehrsführung problematisch geworden und wurde von der Bevölkerung gemieden.” [@lenherr_fortwahrendes_2022, p. 92f.]

::: {layout-ncol=2}

![[Aeschengraben 68](https://osm.org/go/0CdsKSuKY?m=), facing southeast. June 2024.](./media/Aeschenvorstadt68_20240623_143654.jpg){label="fig-aeschengraben68-1"}

![[Aeschengraben 68](https://osm.org/go/0CdsKS3hA?m=), facing southeast. June 2024.](./media/Aeschenvorstadt68_20240623_143532.jpg){label="fig-aeschengraben-2"}

:::

## Current Situation

Kreis 2005, p. 57: "Ironie des Schicksals: Das an der Stelle von Heimatschützern verteidigten Häusern in den Jahren 1954–1957 errichtete 'Drachen-Center' hätte nach dem Willen der staatlichen Denkmalpflege 2005 vor einer mit starken Eingriffen verbundenen Umnutzung geschützt werden sollen. Dieser ehemalige Neubau gilt mittlerweile als exemplarischer Bau eines frühen City-Hauses [...]"

- some regulations still valid
- some regulations were retracted starting in the 19xxxs
- today's use of the urban space gained by facade relocation

  - bars, trees, parking, non-lieu?

::: {layout-ncol=2}

![[Feldbergstr. 63](https://osm.org/go/0CdsPBYbL?m=), facing northwest. June 2024.](./media/Feldbergstr63_20240623_131322.jpg)

![[Klybeckstr. 42](https://osm.org/go/0CdsN6Vrv?m=), facing southeast. June 2024.](./media/Klybeckstr42_20240623_130717.jpg)

:::

- the role of men/politicians in this? nature metaphors: "städtebauliche[] Erosion" [@fingerhuth_bauten_1988, p, 188]
- "it has to be completed now": “Das Korrektionswerk war jetzt so weit fortgeschritten, daß es auch zu Ende geführt werden mußte, obschon sich inzwischen gezeigt hatte, daß der Zweibahnverkehr nicht nötig war und die Verbreiterung noch lange hätte unterbleiben können. Sie bekam indessen erneut einen Sinn durch das 1989 vorgelegte Projekt, ein drittes Tramgeleise in die Vorstadt zu legen.” [@suter_von_1991, p. 25]
- (ggfs.:) looking back from 1998: [@schneider-sliwa_strategien_1998]
- Rückblickend Kreis 2005 p. 53: "Mit fragwürdigen Mitteln wurden fragwürdige Ziele verfolgt."

- parallels on current railway planning: https://www.bazonline.ch/gopferdeckel-herr-de-meuron-382416996850
  - traffic situation has not really changed

# Nauentunnel/Heuwaage-Viadukt

Life after the Correction Plans

![[Heuwaage-Viadukt](https://osm.org/go/0CdsIwfLx--?m=), facing southeast. June 2024.](./media/HeuwaageViadukt_20240610_213622.jpg){width=50%}
